Race Science in Mainstream Psychology, 1960-2010

### 2025-01-03 paper/paper.qmd
table S2

---
title: "Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: a
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    orcid: "0000-0001-7945-4416"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: a
    orcid: "0000-0002-3066-2932"

@@

format: 
  elsevier-pdf:  # https://github.com/quarto-journals/elsevier
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    journal:
      name: "NA"
      cite-style: "authoryear"
      layout: "onecolumn"
    include-in-header: 
      text: |
        \usepackage{lineno}
        % \linenumbers

crossref:
  fig-prefix: ""

@@

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisAppropriationBiological2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'[sic]dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].

@@

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that MQ might have served as an "incubator" for race science discourse, with ideas originating in MQ before appearing within the scientific mainstream.  Establishing venues for favorable research is another key element of the "tobacco strategy" for manufacturing doubt [@FernandezPintoKnowBetterNot2017].  Specifically, we hypothesized that scientometric and text mining analyses would show race science discourse appearing first and more prominently in MQ, and appearing in mainstream publications only subsequently.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science discourse, based on intelligence research and originating outside of MQ.

@@

NLP-extracted noun phrases offer a number of advantages over the more traditional unigram ("single word") terms.  First, noun phrase extraction removes many standard stopwords (articles, common verbs, numbers) without relying on a fixed, a priori stopword list.  Phrases can be more informative than single terms, for example, distinguishing "intelligence test" from "hypothesis test."  While simple n-gram extraction will include numerous phrases that are not especially meaningful.  Consider the sentence "Since our first analyses of feeding patterns in rats, we had been using a criterion of 40 minutes (Le Magnen & Tallon 1963; 1966)" [@SclafaniCorrelationCausationStudy1981].  Bigrams such as "since our" and "criterion 40" will likely be discarded in vocabulary selection, but significantly increase the computational cost of vocabulary selection.  Noun phrase extraction is therefore more efficient.

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal}

Figure @fig-focal focuses on three topics identified in one of the 24 fitted topic models; see figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models, and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models; we discuss the robustness of our findings across all models below.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or these were not available for the corpus.

@@

BG published very few articles in any race-and-intelligence topic identified by any of the 24 topic models.  But, among both (other) academics and the general public, the field of behavior genetics is strongly associated with race science, and specifically race-and-intelligence research. Panofsky [@PanofskyMisbehavingScienceControversy2014] argues that, prior to Jensen's 1969 paper, behavior genetics emphasized the study of non-human animals and intentionally avoided associations with eugenics and public controversy more generally. In response to Jensen, critics such as Lewontin offered broad critiques of behavior genetics as such, and behavior geneticists in turn adopted a radical conception of academic freedom (without any sense of responsibility for the social implications of academic research) and a siege or wartime mentality, as illustrated by Sandra Scarr's 1986 presidential address to the Behavior Genetics Association (BGA) [@ScarrThreeCheersBehavior1987].  Scarr's address coincided with the period between 1970-1990 when BG published articles (including Scarr's address itself) in the race science discourse topic (24) of the medium vocabulary, $k = 40$ model.

However, the topic model analysis suggests that, by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation was supported by a supplemental analysis that focused on BG specifically; see @sec-bg.  For example, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marxist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities were caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. Behavior genetics may have been prominent in promoting race science in the past, but appears to have been less receptive to such ideas in more recent history.  Instead, our results suggest that other subdisciplines of psychology — perhaps especially cognitive psychology — have provided the primary venue for mainstreaming race science.

@@

@@

For $k=5, 10$ the distributions are quite noisy, difficult to interpret, and generally don't form very strong visual clusters.  As $k$ increases from 20 through 40, coherent bands start to appear for journal-distinctive topics in MQ, BG, Int and, to a lesser degree, PID.  Note that the distinctive topics for MQ and Int do not obviously coincide.  Above about $k=50$, the models appear to identify more fine-grained topics and the coherent bands fade.

>>>>

---
title: "Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "D.J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: a
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    orcid: "0000-0001-7945-4416"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: a
    orcid: "0000-0002-3066-2932"

@@

format: 
  elsevier-pdf:  # https://github.com/quarto-journals/elsevier
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    journal:
      name: "NA"
      cite-style: "authoryear"
      layout: "onecolumn"
    include-in-header: 
      text: |
        \usepackage{lineno}
        \linenumbers

crossref:
  fig-prefix: ""
  tbl-prefix: ""

@@

Histories of scientific racism in the twentieth century have often emphasized the Pioneer Fund and the parascholarly journal *Mankind Quarterly* (MQ) [
@BarkanRetreatScientificRacism1992;
@MehlerFoundationFascismNew1989;
@WinstonScienceServiceFar1998;
@TuckerFundingScientificRacism2002;
@SchafferScientificRacismAgain2007;
@SainiSuperiorReturnRace2019;
@WinstonScientificRacismNorth2020;
@AdamsMisAppropriationBiological2021;
@SainiDraperMillionsPhilanthropic2022].  Pioneer was formed in 1937, during the waning years of the eugenics movement in North America [@BarkanRetreatScientificRacism1992; @TuckerFundingScientificRacism2002], with the aim of "support[ing] academic research and the \'dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 p.21, quoting Laughlin, unbalanced quotation marks in Mehler].  MQ was founded in 1960 by biologist R. Ruggles Gates (1882-1962), psychologist Henry Garrett (1894-1973), and non-academic anthropologist G. Robert Gayre (self-styled as "Gayre of Gayre and Nigg"; 1907-1996).  Gates' professional status had risen and fallen with eugenics and the explicit scientific racism of the 1920s, and by the end of the Second World War he was thoroughly marginalized [@WinstonScienceServiceFar1998].  In contrast, Garrett had been president of the American Psychological Association in 1946 and chair of Psychology at Columbia from 1941 to 1955 [@WinstonScienceServiceFar1998].

@@

Scholars in the agnotology ("theory of ignorance") literature have examined the ways that powerful economic interests, especially the fossil fuel and tobacco industries, have "manufactured doubt," creating a inaccurate public perception of scientific uncertainty to delay public protective regulation [@OreskesMerchantsDoubtHow2011; @FernandezPintoKnowBetterNot2017; @KouranyScienceProductionIgnorance2020]. Following @OreskesMerchantsDoubtHow2011, @FernandezPintoKnowBetterNot2017 identified five "core strategies or mechanisms" of the "tobacco strategy," showing how this strategy was used to manufacture doubt surrounding climate change, pharmaceuticals, and the causes of the 2008 financial crisis.

Race science has not been heavily studied using an agnotology framework [though see @MillsRacialContract2014 and @KouranyMightScientificIgnorance2020].  But the qualitative historiography suggests that actors such as the Pioneer Fund have deployed a version of the tobacco strategy to promote misleading public perceptions of scientific uncertainty about (or even scientific support for) racial hierarchy. (Indeed, the origins of the "tobacco strategy" are often dated to a 1954 advertising campaign by the tobacco industry, 17 years after Pioneer was founded. Scientific racists and the eugenics movement might have been deploying the "tobacco strategy" for decades before the tobacco industry.)

The Pioneer Fund itself clearly exemplifies the strategies of supporting friendly research; and figures such as Shockley and Jensen exemplify the strategy of recruiting high-profile, distinguished scientists *[Jackson? on funding Shockley]*. Using agent-based models, *[cailin etc.]* have shown that *[st]*.

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "boundary."  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be "incubated," protected from the destructive critical scrutiny they would receive in the mainstream. Boundary venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research.

MQ does not appear to have ever been considered a mainstream journal; it has always existed firmly on the margins. If MQ provided the "safe space" for "incubating" race science, away from mainstream criticism, then boundary venues would be necessary to disseminate race science into the scientific mainstream. We therefore hypothesized (H1) that scientometric and text mining analyses would show race science discourse appearing first and most prominently in MQ, and appearing in mainstream publications only subsequently.

Specifically, among both (other) academics and the general public, the field of behavior genetics is strongly associated with race science [@PanofskyMisbehavingScienceControversy2014]. We therefore hypothesized (H2) that the journal *Behavior Genetics* (BG) would be a major boundary venue for the dissemination of race science discourse.

Neither of these hypotheses were supported. Instead, mainstream psychology journals provided a venue for a distinct form of race science discourse, based on intelligence research and originating outside of MQ. And BG has not been a major venue for either form of race science discourse.

@@

NLP-extracted noun phrases offer a number of advantages over the more traditional unigram ("single word") terms.  First, noun phrase extraction removes many standard stopwords (articles, common verbs, numbers) without relying on a fixed, a priori stopword list.  Phrases can be more informative than single terms, for example, distinguishing "intelligence test" from "hypothesis test."  While simple n-gram extraction will include numerous phrases that are not especially meaningful.  Consider the sentence "Since our first analyses of feeding patterns in rats, we had been using a criterion of 40 minutes (Le Magnen & Tallon 1963; 1966)" [@SclafaniCorrelationCausationStudy1981].  Bigrams such as "since our" and "criterion 40" will likely be discarded in vocabulary selection, but significantly increase the computational cost of vocabulary selection.  Noun phrase extraction is therefore more efficient.  For examples of previous research using noun phrases, see @HicksImpactingCapabilitiesConceptual2018; @HicksProductivityInterdisciplinaryImpacts2021.

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic; text version in table @tbl-silge.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal}

Figure @fig-focal focuses on three topics identified in one of the 24 fitted topic models; see table @tbl-silge for a table version of the Silge plot in @fig-focal; figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models, and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models; we discuss the robustness of our findings across all models below.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or these were not available for the corpus.

@@

BG published very few articles in any race-and-intelligence topic identified by any of the 24 topic models.  But, among both (other) academics and the general public, the field of behavior genetics is strongly associated with race science, and specifically race-and-intelligence research. Panofsky [@PanofskyMisbehavingScienceControversy2014] argues that, prior to Jensen's 1969 paper, behavior genetics emphasized the study of non-human animals and intentionally avoided associations with eugenics and public controversy more generally. In response to Jensen, critics such as Lewontin offered broad critiques of behavior genetics as such, and behavior geneticists in turn adopted a radical conception of academic freedom (without any sense of responsibility for the social implications of academic research) along with a siege or wartime mentality, as illustrated by Sandra Scarr's 1986 presidential address to the Behavior Genetics Association (BGA) [@ScarrThreeCheersBehavior1987].  Scarr's address coincided with the period between 1970-1990 when BG published articles (including Scarr's address itself) in the race science discourse topic (24) of the medium vocabulary, $k = 40$ model.

However, the topic model analysis suggests that, by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation was supported by a supplemental analysis that focused on BG specifically; see @sec-bg.  For example, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marxist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities were caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. Behavior genetics may have been prominent in promoting race science in the past, but appear to have been less receptive to such ideas in more recent history.  Instead, our results suggest that other subdisciplines of psychology — perhaps especially cognitive psychology — have provided the primary venue for mainstreaming race science.

@@

```{r}
#| tbl-cap: "Top 15 terms, by beta value, from selected topics of the medium vocabulary, $k = 40$ model"
#| label: tbl-silge
#| echo: false
library(gt)

readRDS(file.path('..', 'out', '08_silge_gt.Rds')) |>
    # cols_width(everything() ~ pct(100/6)) |>
    tab_options(table.font.size = 10)
```

@@

For $k=5, 10$ the distributions are quite noisy, difficult to interpret, and generally don't form very strong visual clusters.  For moderate values of $k$, 20 through 40, coherent bands start to appear for journal-distinctive topics in MQ, BG, Int and, to a lesser degree, PID.  Note that the distinctive topics for MQ and Int do not obviously coincide.  At a low level, these journal-distinctive topics reflect different distributions of noun phrases, and prima facie are likely explained by the relatively narrow scope of these four journals; PR is a generalist journal covering all of psychology and BBS is an interdisciplinary cognitive science journal.  Above about $k=50$, the models appear to identify more fine-grained topics, the coherent bands fade, and the topics do not obviously distinguish different journals.
